---
title: "Figure 1"
output: html_document
---

## Setup - load packages and folder structure
```{r, message=FALSE}
rm(list = ls())

# load required packages
library(data.table)
library(ggplot2)
library(rgdal)
library(sf)
library(dplyr)
library(tidyr)
library(cowplot)
library(MetBrewer)
library(scales)
library(readr)
library(stringr)
library(here)

# set up folder paths
project.folder = paste0(here::here(), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

## Prepare map structure
```{r}
theme_map = function(base_size = 10, base_family = "") {
  require(grid)
  theme_bw(base_size = base_size, base_family = base_family) %+replace%
    theme(
      axis.line = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      panel.background = element_blank(),
      panel.border = element_blank(),
      panel.grid = element_blank(),
      panel.spacing = unit(0, "lines"),
      plot.background = element_blank(),
      legend.position = 'bottom'
    )
}
```

## Load processed data (from b_00_native_land_explore.Rmd)
```{r}
shapefile.data <- readRDS(paste0(project.folder, "data/processed/shapefile_data_all.rds"))
us.main <- readRDS(paste0(project.folder, "data/processed/us_main_shapefile.rds"))
USA.df <- readRDS(paste0(project.folder, "data/processed/usa_map_data.rds"))

# Ensure consistent types
shapefile.data$GEOID <- as.character(shapefile.data$GEOID)
USA.df$native_id <- str_pad(as.character(USA.df$native_id), width = 7, side = "left", pad = "0")
```

## Load WBGT regression file
```{r}
dat.wbgt.summarised.regression <- readRDS(
  paste0(wbgt.folder, 'weighted_area_raster_native_wbgtmax_daily_',
         start_year_wbgt, '_', end_year_wbgt, '_regression_analysis.rds')
) %>%
  mutate(native_id = str_pad(as.character(native_id), width = 7, side = "left", pad = "0"))
```

## Load WBGT over time file
```{r}
dat.wbgt.summarised <- readRDS(
  paste0(wbgt.folder, 'weighted_area_raster_native_wbgtmax_daily_',
         start_year_wbgt, '_', end_year_wbgt, '_over_time.rds')
) %>%
  mutate(
    native_id = str_pad(as.character(native_id), width = 7, side = "left", pad = "0"),
    year = as.numeric(as.character(year))
  )
```

## Prepare data for Panel A: Territory bar chart (now self-contained)
```{r}
# Build the merged data that was previously created in b_02
dat.wbgt.summarised.merged.weighted.native.more.info <- dat.wbgt.summarised %>%
  left_join(
    shapefile.data %>% select(GEOID, NAME, NAMELSAD, STATE_NAME, FUNCSTAT, POP_2020),
    by = c('native_id' = 'GEOID')
  ) %>%
  filter(!is.na(STATE_NAME), !is.na(POP_2020), POP_2020 > 0)

dat.territory <- dat.wbgt.summarised.merged.weighted.native.more.info

## Calculate population-weighted values by territory and year
dat.territory.year <- dat.territory %>%
  group_by(native_id, NAME, NAMELSAD, STATE_NAME, year) %>%
  summarise(
    wbgt_28 = weighted.mean(wbgt_28, POP_2020, na.rm = TRUE),
    POPULATION = sum(POP_2020),
    .groups = "drop"
  ) %>%
  mutate(wbgt_28_pop = wbgt_28 * POPULATION)

## Take 5-year average for each territory
dat.territory.5year <- dat.territory.year %>%
  filter(year %in% years_wbgtmax[c((length(years_wbgtmax)-4):length(years_wbgtmax))]) %>%
  group_by(native_id, NAME, NAMELSAD, STATE_NAME) %>%
  summarise(
    wbgt_28_pop_mean = mean(wbgt_28_pop),
    .groups = "drop"
  )

## Count territories per state and create state labels
state_territory_counts <- dat.territory.5year %>%
  group_by(STATE_NAME) %>%
  summarise(
    n_territories = n_distinct(native_id),
    total_person_days = sum(wbgt_28_pop_mean),
    .groups = "drop"
  ) %>%
  mutate(STATE_LABEL = paste0(STATE_NAME, " (", n_territories, ")"))

## Calculate rank and merge
dat.rank <- state_territory_counts %>%
  arrange(total_person_days) %>%
  mutate(rank = row_number())

## Plot data preparation - use GEOID-based unique labeling to handle duplicate names
dat.plot <- dat.territory.5year %>%
  left_join(state_territory_counts, by = "STATE_NAME") %>%
  left_join(dat.rank %>% select(STATE_NAME, rank), by = "STATE_NAME")

top_n_count <- 100 # of Territories included

# Use NAMELSAD for unique identification (includes type suffix like "Reservation", "Colony", etc.)
tribe_totals <- dat.plot %>%
  group_by(native_id, NAME, NAMELSAD) %>%
  summarise(total = sum(wbgt_28_pop_mean), .groups = "drop")

top_tribes <- tribe_totals %>%
  arrange(desc(total)) %>%
  slice_head(n = top_n_count)

dat.top <- dat.plot %>% filter(native_id %in% top_tribes$native_id)
```

## Load US state shapefile for mapping
```{r}
us <- readOGR(dsn = paste0(data.folder, "shapefiles/states"), layer = "states", verbose = FALSE)

us_aea <- spTransform(
  us,
  CRS("+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0 +a=6370997 +b=6370997 +units=m +no_defs")
)
us_aea@data$id <- rownames(us_aea@data)
us_aea <- us_aea[us_aea$STATE_FIPS %in% states_included, ]
us_aea <- spTransform(
  us_aea,
  CRS("+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0 +a=6370997 +b=6370997 +units=m +no_defs")
)
us_aea <- fortify(us_aea)
```

## Create Panel A: Territory stacked bar chart
```{r}
# Create unique label for each territory (handles duplicate NAMEs across states)
dat.top <- dat.top %>%
  mutate(territory_label = ifelse(
    duplicated(NAME) | duplicated(NAME, fromLast = TRUE),
    paste0(NAMELSAD, " (", STATE_NAME, ")"),
    NAMELSAD
  ))

plot.territory.bar.chart <-
  ggplot(dat.top,
         aes(x = reorder(territory_label, wbgt_28_pop_mean, FUN = sum),
             y = wbgt_28_pop_mean,
             fill = STATE_NAME)) +
  geom_col(width = .75, colour = "white", linewidth = .125, position = "stack") +
  coord_flip() +
  labs(
    y = paste0('Mean annual dangerous hot-humid person-days\n',
               end_year_wbgt - 4, '-', end_year_wbgt),
    fill = "State",
    x = paste0('Top 100 Territories by Hazardous Heat Exposure')
  ) +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0.02, 0.15))) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.major.x = element_line(colour = "gray95", linewidth = .2),
    panel.grid.major.y = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background  = element_rect(fill = "white"),
    axis.text.x = element_text(size = 9, margin = margin(t = 8)),
    axis.text.y = element_text(size = 5, margin = margin(r = 8)),
    axis.title.x = element_text(margin = margin(t = 12), size = 10),
    axis.title.y = element_text(margin = margin(r = 12), size = 10),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.box = "horizontal"
  )
```

## Prepare summary maps
```{r}
USA.df.merged <- USA.df %>%
  mutate(native_id = as.character(native_id)) %>%
  left_join(
    shapefile.data %>% select(GEOID, NAME, STATE_NAME, FUNCSTAT, POP_2020),
    by = c("native_id" = "GEOID")
  ) %>%
  filter(!is.na(POP_2020), POP_2020 > 0) %>%
  left_join(dat.wbgt.summarised.regression, by = 'native_id') %>%
  arrange(id, order)
```

## Prepare colors and legends of maps
```{r}
legend_create_count_last_5_years = function(threshold) {
  paste0('Mean annual number of dangerous hot-humid\ndays (WBGT > ', threshold,
         '\u00B0C) during ', (end_year_wbgt - 4), '-', end_year_wbgt)
}

plot_last_5_years = function(threshold_chosen, legend_chosen = 0) {
  fill_col <- paste0('wbgt_', threshold_chosen, '_', (end_year_wbgt - 4), '_', end_year_wbgt)
  max_val <- max(USA.df.merged[[fill_col]], na.rm = TRUE)

  p = ggplot() +
    geom_polygon(data = us_aea, aes(x = long, y = lat, group = group),
                 fill = alpha('grey35', 1),
                 color = alpha('grey15', 0.9),
                 size = 0.25) +
    geom_polygon(data = USA.df.merged,
                 aes(x = long, y = lat, group = group,
                     fill = .data[[fill_col]]),
                 color = scales::alpha("white", 0.01),
                 size = 0.01) +
    scale_fill_gradientn(colors = rev(MetBrewer::met.brewer('Tam')),
                         limits = c(0, max_val),
                         na.value = "grey35") +
    coord_fixed() + xlab('') + ylab('') + theme_bw() +
    theme(panel.grid.major = element_blank(),
          axis.text.x = element_text(angle = 90),
          axis.ticks.x = element_blank(),
          legend.text = element_text(size = 5),
          panel.grid.minor = element_blank(),
          axis.line = element_line(colour = "black"),
          panel.border = element_rect(colour = "black"),
          strip.background = element_blank(),
          legend.justification = 'center',
          legend.position = 'bottom',
          legend.background = element_rect(fill = "white", size = .5, linetype = "dotted")) +
    theme_map()

  if (legend_chosen == 0) {
    p = p + guides(fill = "none")
  }
  if (legend_chosen == 1) {
    p = p +
      guides(fill = guide_colorbar(direction = "horizontal",
                                   title.position = "left",
                                   barwidth = 10,
                                   barheight = 1,
                                   title.vjust = 0.8,
                                   title = legend_create_count_last_5_years(threshold_chosen)))
  }

  return(p)
}

plot.count.last.5.years.28 = plot_last_5_years(28, 1)
```

## Plot and save
```{r}
pdf(paste0(figures.folder, 'Figure_1.pdf'),
    paper = 'a4r',
    width = 0,
    height = 0)
print(plot_grid(plot.territory.bar.chart,
                plot.count.last.5.years.28,
                label_size = 14,
                label_x = -0.02,
                label_y = 1,
                labels = c('(a)', '(b)')))
dev.off()

plot_grid(plot.territory.bar.chart,
          plot.count.last.5.years.28,
          label_size = 14,
          label_x = -0.02,
          label_y = 1,
          labels = c('(a)', '(b)'))
```
